---
title: "Benchmarking against a plain R approach"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE
)
```

A critical reader suggest that, given that `mdl::mtrx()` does quite a bit less than `model.matrix()`, it's not really suprising that it would run more quickly than `model.matrix()`. How fast could we make an analogue to `mdl::mtrx()` in "plain R" (i.e. without rust)? Let's see.

```{r setup}
library(mdl)
library(bench)
library(ggplot2)
library(dplyr)
```

## A plain R alternative

`mdl::mtrx()` is mostly just a glorified `as.matrix()` method for data frames that makes dummy variables out of factors, characters, and logicals. Let's see if we mock up a plain R alternative, `mdl_mtrx()`, that's just as fast.

First, we'll write a helper to apply to factors to convert them to dummy variables, loosely based on `hardhat::fct_encode_one_hot()`.

```{r}
x <- factor(sample(letters[1:3], 100, replace = TRUE))

fct_encode_dummy <- function(x) {
  row_names <- names(x)
  col_names <- levels(x)
  col_names <- col_names[-1]
  dim_names <- list(row_names, col_names)
  n_cols <- length(col_names)
  n_rows <- length(x)
  x <- unclass(x)
  out <- matrix(0L, nrow = n_rows, ncol = n_cols, dimnames = dim_names)
  loc <- cbind(row = seq_len(n_rows), col = x - 1)
  out[loc] <- 1L
  out
}
```

It's hard to beat R at converting characters to factors, so we'll convert character vectors to factors and then to dummy variables via `fct_encode_dummy()`. The rest of the variable types we'll test here are one-liners.

```{r}
to_numeric <- function(x) {
  switch(
    class(x),
    numeric = x,
    integer = x,
    character = fct_encode_dummy(as.factor(x)),
    factor = fct_encode_dummy(x),
    logical = x
  )
}
  
mdl_mtrx <- function(formula, data) {
  predictors <- mdl:::predictors(formula, data)
  
  cols <- lapply(data[predictors], to_numeric)
  do.call(cbind, cols)
}
```

We'll use a data frame with a variety of types to benchmark in this article. Wrapping in a quick function:

```{r}
create_data_frame <- function(n_rows) {
  data.frame(
    outcome = runif(n_rows),
    pred_numeric = runif(n_rows),
    pred_integer = sample(c(0L, 1L), n_rows, replace = TRUE),
    pred_logical = sample(c(TRUE, FALSE), n_rows, replace = TRUE),
    pred_factor_2 = factor(sample(letters[1:2], n_rows, replace = TRUE)),
    pred_factor_3 = factor(sample(letters[1:3], n_rows, replace = TRUE)),
    pred_character_2 = sample(letters[1:2], n_rows, replace = TRUE),
    pred_character_3 = sample(letters[1:3], n_rows, replace = TRUE)
  )
}

d <- create_data_frame(5)

d
```
Passing each of those data types to make sure our new function `mdl_mtrx()` does what we want in the simplest case:

```{r}
mdl_mtrx(outcome ~ ., d)
```

Okay, so, no type checking, plenty of issues with edge cases, lean and mean R function. Let's check the timings out, first on a very small dataset:

```{r}
d <- create_data_frame(30)

bench::mark(
  mdl_mtrx = mdl_mtrx(outcome ~ ., d),
  `mdl::mtrx` = mdl::mtrx(outcome ~ ., d),
  model.matrix = model.matrix(outcome ~ ., d),
  check = FALSE
)
```

Not bad, scrappy little feller! Now, on a few more reasonably sized datasets:

```{r, cache = TRUE}
res <- 
  bench::press(
    n_rows = 10^seq(2:8), 
    {
      d <- create_data_frame(n_rows)
      bench::mark(
        mdl_mtrx = mdl_mtrx(outcome ~ ., d),
        `mdl::mtrx` = mdl::mtrx(outcome ~ ., d),
        model.matrix = model.matrix(outcome ~ ., d),
        check = FALSE
      )
    }
  )

ggplot(res) +
  aes(x = n_rows, y = median, col = expression) +
  geom_point() +
  scale_x_log10()
```

If the log scale is tripping you up, here's a slice of data for the largest numbers of rows:

```{r}
res %>% 
  select(expression, median, mem_alloc) %>%
  tail(3)
```

```{r}
timings <- res %>% 
  tail(3) %>%
  pull(median) %>%
  as.numeric()
```

In this timing, `mdl::mtrx()` was `r round(1 - (timings[1] / timings[2]), 3)`% faster than the plain R approach, and `r round(1 - (timings[3] / timings[2]), 3)`% slower than `model.matrix()`.

